Fusing Filters with Integer Linear Programming

### 2014-05-16 2014betterfusionforfilters/section/04-ILP.tex
fix possible
remove normalize2 code from transducer 3.6, get trans defn to fit without break
dep graph
consistent commas in x_{i,j}

\begin{figure}
\begin{tabbing}
MMMMM       \= M  \= \kill
$nodes$     \> @:@ \> $\function \to V$              \\
$edges$     \> @:@ \> $\function \to E$              \\
$edge$      \> @:@ \> $\{bind\} \times bind \to E$ 
\\[1ex]
$nodes(bs)$ \> $= \{(name(b), \iiter_{\Gamma,C}(b)) | b \in bs\}$       
\\[2ex]

$edges(bs)$ \> $= \bigcup_{b \in bs}edge(bs, b)$    
\\[1ex]
MM             \= M \= \kill
$edge(bs, out = @fold@~f~in)$ \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @map@~f~in)$  \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @filter@~f~in)$  \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @gather@~data~indices)$  \\
    \> $=$    \> $\{(out,data, \fusionpreventing) \} \cup \{inedge(bs, out, indices) \}$       \\
$edge(bs, out = @cross@~a~b)$            \\
    \> $=$    \> $\{inedge(bs, out, a) \}           \cup      \{(out, b, \fusionpreventing) \}$ \\
$edge(bs, outs = @external@~ins)$  \\
    \> $=$    \> $\{(outs,i, \fusionpreventing) | i \in ins \}$ 
\\[1ex]
$inedge(bs,to,\from)$ \\
    \> $|$ \> $(\from = @fold@~f~s) \in bs$     \\
    \> $=$ \> $(to, \from, \fusionpreventing)$  \\
    \> $|$ \> $(outs = @external@ \ldots) \in bs     \wedge \from \in outs$     \\
    \> $=$ \> $(to, outs, \fusionpreventing)$  \\
    \> $|$ \> $otherwise$                      \\
    \> $=$ \> $(to, \from, \fusible)$
\end{tabbing}

@@

\eject
% -----------------------------------------------------------------------------
\subsection{ILP Variables}
After generating the dependency graph, the next step is to produce a set of linear constraints from this graph. The variables involved in these constraints are split into three groups:
\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$x$   \> @:@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$
\end{tabbing}
For each pair of nodes with indices $i$ and $j$ we use a boolean variable $x_{ij}$ which indicates whether those two nodes are fused. We use $x_{ij} = 0$ when the nodes are fused and $x_{ij} = 1$ when they are not. Using $0$ for the fused case means that the objective function can be a weighted function of the $x_{ij}$ variables, and minimizing it tends to increase the number of nodes that are fused. The values of these variables are used to construct the final clustering, such that $\forall i,j.\ x_{ij} = 0 \iff cluster(i) = cluster(j)$.
\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$\pi$ \> @:@  \> $node$             \> $\to$ \> $\mathbb{R}$
\end{tabbing}
The second group of variables is used to ensure that the clustering is acyclic. This means that for each node in the graph, the dependencies of that node can be executed before the node itself. For each node $i$, we associate a real $\pi_i$ such that every node $j$ that depends on $i$ we have $\pi_j > \pi_i$. Our linear constraints will ensure that if two nodes are fused into the same cluster then their $\pi$ values will be identical --- though nodes in different clusters can also have the same $\pi$ value.

Here is an example of a cyclic clustering:
\begin{code}
  cycle xs  = let ys  = map (+1) xs     (C1)
                  sum = fold ys         (C2)
                  zs  = map (+sum) ys   (C1)
              in  zs
\end{code}
There is no fusion-preventing edge directly between the @xs@ and @zs@ bindings, but there is a fusion-preventing edge between @sum@ and @zs@. If the @xs@ and @zs@ bindings were in the same cluster @C1@ and @sum@ was in cluster @C2@, there would be a dependency cycle between @C1@ and @C2@, and neither could be executed before the other.
\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$c$   \> @:@  \> $node$             \> $\to$ \> $\mathbb{B}$
\end{tabbing}
The final group of variables is used to help define the cost model encoded by the objective function. Each node is assigned a variable $c_i$ that indicates whether the array the associated binding produces is \emph{fully contracted}. When an array is fully contracted it means that all consumers of that array are fused into the same cluster, so we have $c_i = 0 \iff \forall (i',j) \in E.\ i = i' \implies x_{ij} = 0$. In the final program, each successive element of a fully contracted array can be stored in a scalar register, rather than requiring an array register or memory storage.

@@

% -------------------------------------
\paragraph{Array contraction} The final group of constraints gives meaning to the $c$ variables that we use to represent whether an array is fully contracted:
\begin{tabbing}
MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
        \> $x_{ij}$    \> $\le$ \> $c_i$  \> \> \\
        \> (for all edges from i)
\end{tabbing}
Recall that an array is fully contracted when all of the consumers are fused with the nodes that produces it, which means that the array does not need to be fully materialized in memory. As per Darte's work on array contraction~\cite{darte2002contraction}, we define a variable $c_i$ for each array, and the constraint above ensures that $c_i = 0$ only if $\forall (i',j) \in E.\ i = i' \implies x_{ij} = 0$. By minimizing $c_i$ in the objective function, we favor solutions that reduce the number of intermediate arrays.

@@

To find the objective function, note that fusing loops can have three main benefits:
\begin{itemize}
\item
reducing memory traffic, such as multiple loops reading from the same array;
\item
removing intermediate arrays, thus reducing the amount of memory required;
\item
and reducing loop overhead, such as when two loops operate on different arrays of the same size.
\end{itemize}
However, these benefits cannot be considered in isolation; for example, fusing two loops to reduce loop overhead may remove potential fusion opportunities that reduce memory traffic.
When operating on large arrays that do not fit in cache, memory traffic dominates execution time.
An excessive number of intermediate arrays can also cause issues if all are live in memory at once, potentially leading to \emph{thrashing}.
The benefits of removing loop overhead are least of all; it should be performed if possible, but must never remove opportunities for other fusion. \TODO{Refer to example code that demonstrates these different opportunities.}

@@

\begin{tabbing}
MMMMM   \= MMMM \= M \= \kill
Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{ij} \cdot x_{ij}$   \\
           \> \> \> (memory traffic and loop overhead)         \\
           \> $+$ \> $\Sigma_{i \in V} N \cdot c_i$  \\
           \> \> \> (removing intermediate arrays)         \\
Subject to \> \ldots                                \\
Where      \>                                       \\
           \> $W_{ij} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
           \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
           \> $W_{ij} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
           \> \> \> ($i$ and $j$ share an input array)                                         \\
           \> $W_{ij} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
           \> \> \> (the only benefit is loop overhead)                                        \\
           \\
           \> $N = |V|$
\end{tabbing}

% -----------------------------------------------------------------------------
\subsection{Transitivity of Clustering}
It may seem that we could generate far fewer constraints and rely on transitivity of clustering equality $x_{ij}$.
This means that for each pair $i,j$ we would only generate an $x_{ij}$ and constraints, if $i$ or $j$ is a @filter@, or there is an edge between $i$ and $j$.
This would simplify the integer linear program, but would also modify the objective function not to \emph{minimise} on these removed $x_{ij}$.
The objective function of this simplified version would not count the benefits of fusing two non-connected nodes, and would not reduce memory traffic as effectively.

@@

\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> $\Sigma_{(i,j) \in E} W_{ij} \cdot x_{ij} + \Sigma_{i \in V} N \cdot c_i$  \\
           \> (if $possible(i,j)$)         \\
Subject to \\
           \> $-N \cdot x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{ij}$ \\
           \> (if $possible(i,j) \wedge (i,j) \not\in E \wedge (j,i) \not\in E$)            \\
\\
           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{ij}$ \\
           \> (if $possible(i,j) \wedge (i,j,\fusible) \in E$)     \\
\\
           \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
           \> (if $(i,j,\fusionpreventing) \in E$)    \\
\\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (if $possible(i,j)$) \\
           \> $c_{i }$    \> $ = $ \> $ 1 $           \>       \>            \\
           \> (if $\neg possible(i,j)$) \\
           \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  \\
           \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge possible'(i,j)$) \\

\\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge \neg possible'(i,j)$) \\
MMMMM   \= MMMM \= M \= \kill
Where      \>                                       \\
           \> $W_{ij} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
           \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
           \> $W_{ij} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
           \> \> \> ($i$ and $j$ share an input array)                                         \\
           \> $W_{ij} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
           \> \> \> (the only benefit is loop overhead)                                        \\
           \\
           \> $N = |V|$
\end{tabbing}

>>>>

\begin{figure}
\begin{tabbing}
MMMMM       \= M  \= \kill
$nodes$     \> @:@ \> $\function \to V$              \\
$edges$     \> @:@ \> $\function \to E$              \\
$edge$      \> @:@ \> $\{bind\} \times bind \to E$   \\
$inedge$    \> @:@ \> $\{bind\} \times name \times name \to E$   \\
\\
$nodes(bs)$ \> $= \{(name(b), \iiter_{\Gamma,C}(b)) | b \in bs\}$       \\
\\
$edges(bs)$ \> $= \bigcup_{b \in bs}edge(bs, b)$    \\
\\
MM             \= M \= \kill
$edge(bs, out = @fold@~f~in)$ \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @map@~f~in)$  \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @filter@~f~in)$  \\
    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
$edge(bs, out = @gather@~data~indices)$  \\
    \> $=$    \> $\{(out,data, \fusionpreventing) \} \cup \{inedge(bs, out, indices) \}$       \\
$edge(bs, out = @cross@~a~b)$            \\
    \> $=$    \> $\{inedge(bs, out, a) \}           \cup      \{(out, b, \fusionpreventing) \}$ \\
$edge(bs, outs = @external@~ins)$  \\
    \> $=$    \> $\{(outs,i, \fusionpreventing) | i \in ins \}$ \\
\\
$inedge(bs,to,\from)$ \\
    \> $|$ \> $(\from = @fold@~f~s) \in bs$     \\
    \> $=$ \> $(to, \from, \fusionpreventing)$  \\
    \> $|$ \> $(outs = @external@ \ldots) \in bs     \wedge \from \in outs$     \\
    \> $=$ \> $(to, outs, \fusionpreventing)$  \\
    \> $|$ \> $otherwise$                      \\
    \> $=$ \> $(to, \from, \fusible)$
\end{tabbing}

@@

\eject
% -----------------------------------------------------------------------------
\subsection{ILP Variables}
After generating the dependency graph, the next step is to produce a set of linear constraints from this graph. The variables involved in these constraints are split into three groups:
\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$x$   \> @:@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$
\end{tabbing}
For each pair of nodes with indices $i$ and $j$ we use a boolean variable $x_{i,j}$ which indicates whether those two nodes are fused. We use $x_{i,j} = 0$ when the nodes are fused and $x_{i,j} = 1$ when they are not. Using $0$ for the fused case means that the objective function can be a weighted function of the $x_{i,j}$ variables, and minimizing it tends to increase the number of nodes that are fused. The values of these variables are used to construct the final clustering, such that $\forall i,j.\ x_{i,j} = 0 \iff cluster(i) = cluster(j)$.
\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$\pi$ \> @:@  \> $node$             \> $\to$ \> $\mathbb{R}$
\end{tabbing}
The second group of variables is used to ensure that the clustering is acyclic. This means that for each node in the graph, the dependencies of that node can be executed before the node itself. For each node $i$, we associate a real $\pi_i$ such that every node $j$ that depends on $i$ we have $\pi_j > \pi_i$. Our linear constraints will ensure that if two nodes are fused into the same cluster then their $\pi$ values will be identical --- though nodes in different clusters can also have the same $\pi$ value.

Here is an example of a cyclic clustering:
\begin{code}
  cycle xs  = let ys  = map (+1) xs     (C1)
                  sum = fold ys         (C2)
                  zs  = map (+sum) ys   (C1)
              in  zs
\end{code}
There is no fusion-preventing edge directly between the @xs@ and @zs@ bindings, but there is a fusion-preventing edge between @sum@ and @zs@. If the @xs@ and @zs@ bindings were in the same cluster @C1@ and @sum@ was in cluster @C2@, there would be a dependency cycle between @C1@ and @C2@, and neither could be executed before the other.
\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$c$   \> @:@  \> $node$             \> $\to$ \> $\mathbb{B}$
\end{tabbing}
The final group of variables is used to help define the cost model encoded by the objective function. Each node is assigned a variable $c_i$ that indicates whether the array the associated binding produces is \emph{fully contracted}. When an array is fully contracted it means that all consumers of that array are fused into the same cluster, so we have $c_i = 0 \iff \forall (i',j) \in E.\ i = i' \implies x_{i,j} = 0$. In the final program, each successive element of a fully contracted array can be stored in a scalar register, rather than requiring an array register or memory storage.

@@

% -------------------------------------
\paragraph{Array contraction} The final group of constraints gives meaning to the $c$ variables that we use to represent whether an array is fully contracted:
\begin{tabbing}
MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
        \> $x_{i,j}$    \> $\le$ \> $c_i$  \> \> \\
        \> (for all edges from i)
\end{tabbing}
Recall that an array is fully contracted when all of the consumers are fused with the nodes that produces it, which means that the array does not need to be fully materialized in memory. As per Darte's work on array contraction~\cite{darte2002contraction}, we define a variable $c_i$ for each array, and the constraint above ensures that $c_i = 0$ only if $\forall (i',j) \in E.\ i = i' \implies x_{i,j} = 0$. By minimizing $c_i$ in the objective function, we favor solutions that reduce the number of intermediate arrays.

@@

To find the objective function, note that fusing loops can have three main benefits:
\begin{itemize}
\item
reducing memory traffic, such as multiple loops reading from the same array;
\item
removing intermediate arrays, thus reducing the amount of memory required;
\item
and reducing loop overhead, such as when two loops operate on different arrays of the same size.
\end{itemize}
However, these benefits cannot be considered in isolation; for example, fusing two loops to reduce loop overhead may remove potential fusion opportunities that reduce memory traffic.
When operating on large arrays that do not fit in cache, memory traffic dominates execution time.
An excessive number of intermediate arrays can also cause issues if all are live in memory at once, potentially leading to \emph{thrashing}.
The benefits of removing loop overhead are least of all; it should be performed if possible, but must never remove opportunities for other fusion.
% \TODO{Refer to example code that demonstrates these different opportunities.}

@@

\begin{tabbing}
MMMMM   \= MMMM \= M \= \kill
Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{i,j} \cdot x_{i,j}$   \\
           \> \> \> (memory traffic and loop overhead)         \\
           \> $+$ \> $\Sigma_{i \in V} N \cdot c_i$  \\
           \> \> \> (removing intermediate arrays)         \\
Subject to \> \ldots                                \\
Where      \>                                       \\
           \> $W_{i,j} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
           \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
           \> $W_{i,j} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
           \> \> \> ($i$ and $j$ share an input array)                                         \\
           \> $W_{i,j} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
           \> \> \> (the only benefit is loop overhead)                                        \\
           \\
           \> $N = |V|$
\end{tabbing}

% -----------------------------------------------------------------------------
\subsection{Transitivity of Clustering}
It may seem that we could generate far fewer constraints and rely on transitivity of clustering equality $x_{i,j}$.
This means that for each pair $i,j$ we would only generate an $x_{i,j}$ and constraints, if $i$ or $j$ is a @filter@, or there is an edge between $i$ and $j$.
This would simplify the integer linear program, but would also modify the objective function not to \emph{minimise} on these removed $x_{i,j}$.
The objective function of this simplified version would not count the benefits of fusing two non-connected nodes, and would not reduce memory traffic as effectively.

@@

\begin{tabbing}
MMMMM   \= MMMM \= M \= MMMM \= M \= MMMM \= \kill
Minimise   \> $\Sigma_{(i,j) \in E} W_{i,j} \cdot x_{i,j} + \Sigma_{i \in V} N \cdot c_i$  \\
           \> (if $possible(i,j)$)         \\
Subject to \\
           \> $-N \cdot x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ \\
           \> (if $possible(i,j) \wedge (i,j) \not\in E \wedge (j,i) \not\in E$)            \\
\\
           \>    $x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ \\
           \> (if $possible(i,j) \wedge (i,j,\fusible) \in E$)     \\
\\
           \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
           \> (if $(i,j,\fusionpreventing) \in E$)    \\
\\
           \> $x_{i,j}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (if $possible(i,j)$) \\
           \> $c_{i }$    \> $ = $ \> $ 1 $           \>       \>            \\
           \> (if $\neg possible(i,j)$) \\
           \\
           \> $x_{i,j}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  \\
           \\
           \> $x_{i',i}$   \> $\le$ \> $x_{i,j}$        \>       \>            \\
           \> $x_{j',j}$   \> $\le$ \> $x_{i,j}$        \>       \>            \\
           \> $x_{i',j'}$   \> $\le$ \> $x_{i,j}$        \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge possible'(i,j)$ \\
           \> \> $\wedge~parents(i,j) = \{(i',j')\}$) \\

\\
           \> $x_{i,j}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge \neg possible'(i,j)$) \\
MMMMM   \= MMMM \= M \= \kill
Where      \>                                       \\
           \> $W_{ij} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
           \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
           \> $W_{ij} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
           \> \> \> ($i$ and $j$ share an input array)                                         \\
           \> $W_{ij} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
           \> \> \> (the only benefit is loop overhead)                                        \\
           \\
           \> $N = |V|$
\end{tabbing}
